Make any stupid plot you want

### 2021-01-20 draft/sections/math.tex
monoids, w stars!

\subsubsection{Variables: Fiber Space $F$}
\label{sec:data_fiber}
The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset. For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ U_{temp} = \{x \in \mathbb{R}^+\}$. 
\begin{equation}
    F = U_{temp}
\end{equation}
Then lets add the time variable from GHCN $U_{time} = \{x \in \mathbb({Z}^+ \mid x \geq 1011832\}$ 
\begin{equation}
F =  U_{temp} \times U_{time}
\end{equation}
such that $F$ is the cartesian product of $F_0=U_{temp}$ and $F_1 = U_{time}$. The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_loc = \{y, x \in \mathbb{R}^{2} \mid -90 \leq y \leq 90, -180 \leq x -180\}$: 
\begin{equation}
    F = U_{temp} \times U_{time} \times U_{loc}
\end{equation}
such that one record $r$ in the fiber would have the form $(temperature,\, time,\, (latitude,\, longitude))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
\begin{align}
    F_0 &= (\textit{temperature}, U_{temp}, M_0)\\
    F_1 &= (\textit{time}, U_{time}, M_1)\\
    F_2 &= (\textit{location}, U_{loc}, M_2)
\end{align}
 or more generally (name, type, monoid)
 \begin{equation}
    F_i = (\textit{name}, U_{type}, M_{i})
 \end{equation}
where type is the set of values defined by the datatype as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}. We describe $M$ in more detail in section~\ref{sec:data_monoid}. We can keep growing the fiber
\begin{equation}
F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
\end{equation}
and identify the elements in the record by the variable name. We will often be working with the decoupled fiber $F_i$ rather than the total fiber $F$.

\subsubsection{Measurement Scales: Monoid Actions}
\label{sec:data_monoid}
A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or enlarging a shape. This model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.

\begin{figure}
    \includegraphics[width=\textwidth]{figures/math/rotation_actions.png}
    \label{fig:data_monoid_rotation}
\end{figure}
One such set of transforms is the rotation group showin in figure~\ref{fig:}
One method of classifying the types of operations that can be performed on data are Steven's measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.

% change to Mi 
We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$. A monoid \cite{Monoid2021} $M$ is a set with an associative binary operator $\ast:M \times M\rightarrow M$. A monoid has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$. A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $F$ with an action $\bullet: M\times F \rightarrow F$ with the properties:
\begin{align*}%lost the \ast
    \textbf{associativity}\;& \text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in F, e\in M,  e\bullet x = x 
\end{align*}

For example, the temperature $F_0$ defined in unit \textdegree C is on an interval scale; this means that the monoid actions $M_0$ on $F_0$ are the set of all transformations $tp_i + c$ such that %%generalize out to x in \U_{temp}, c in \R
%%the action of the interval is the + 
%% c is the element that shifts from x1 to x2
\begin{equation}
F_0 = (\textit{temperature}, \{tp_0, \ldots, tp_n\}, \{tp_0 + c_0, \ldots, tp_0 + c_n, \ldots, tp_n + c_0, \ldots, tp_n + c_n\} ) 
\end{equation}

As with the fiber $F$, the total monoid space $M$ is the cartesian product
\begin{equation}
M = M_{0} \times \ldots \times M_{i}\times \ldots \times\ldots M_{n}
\end{equation}
of the monoids $M_{i}$ on each $F_{i}\in F$. Because each $F_i, M_i$ can be independent of the other fibers, we can formally express the constraints on heterogenous multivariate datasets in the same manner as univariate datasets.

>>>>

\subsubsection{Variables: Fiber Space $F$}
\label{sec:data_fiber}
The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset. For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ U_{temp} = \{x \in \mathbb{R}^+\}$. 
\begin{equation}
    F = U_{temp}
\end{equation}
Then lets add the time variable from GHCN $U_{time} = \{x \in \mathbb({Z}^+ \mid x \geq 1011832\}$ 
\begin{equation}
F =  U_{temp} \times U_{time}
\end{equation}
such that $F$ is the cartesian product of $F_0=U_{temp}$ and $F_1 = U_{time}$. The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_{loc} = \{y, x \in \mathbb{R}^{2} \mid -90 \leq y \leq 90, -180 \leq x -180\}$: 
\begin{equation}
    F = U_{temp} \times U_{time} \times U_{loc}
\end{equation}
such that one record $r$ in the fiber would have the form $(temperature,\, time,\, (latitude,\, longitude))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
\begin{align}
    F_0 &= (\textit{temperature},\, U_{temp},\, M_0)\\
    F_1 &= (\textit{time},\, U_{time},\, M_1)\\
    F_2 &= (\textit{location},\, U_{loc},\, M_2)
\end{align}
 or more generally (name, type, monoid action)
 \begin{equation}
    F_i = (\textit{name},\, U_{type},\, M_{i})
 \end{equation}
where type is the set of values defined by the datatype as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}. We describe $M$ in more detail in section~\ref{sec:data_monoid}. We can keep growing the fiber
\begin{equation}
F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
\end{equation}
and identify the elements in the record by the variable name. We will often be working with the decoupled fiber $F_i$ rather than the total fiber $F$.

\subsubsection{Measurement Scales: Monoid Actions}
\label{sec:data_monoid}

\begin{figure}[ht!]
    \includegraphics[width=\textwidth]{figures/math/rotation_actions.png}
    \label{fig:data_monoid_rotation}
    \caption{The set of rotation actions [turn 0, 90, 180] on the star produce a star (closed), include a rotation that does not change the orientation (identity), and can be done in any order to produce the same final orientation (closure); therefore the rotations are a set of monoid actions.}.  
\end{figure}

A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or rotating a shape such as in figure~\ref{fig:data_monoid_rotation}. This model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}. One method of classifying the types of operations that can be performed on data are Steven's measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}; nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.

We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$. A monoid \cite{Monoid2021} $M_i$ is a set with an associative binary operator $\ast:M_i \times M_i\rightarrow M_i$. A monoid has an identity element $e\in M_i$ such that $e\ast a= a \ast e = a$ for all $a \in M_i$. A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M_i$ is a set $U_{type}$ with an action $\bullet: M\times U_{type} \rightarrow U_{type}$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } m,t \in M \text{ and } x\in U_{type}, m\bullet(t\bullet x) = (m\ast t) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in U_{type}, e\in M,  e\bullet x = x 
\end{align*}

For example, the temperature $F_0$ defined in unit \textdegree C is on an interval scale; this means that the monoid actions $M_0$ on $U_{temp}$ are the set of transformations $\{x + c \in U_{temp} \mid x,c \in U_temp\}$ where $c$ is a valid interval between two temperatures $c=x_i-x_j$.

As with the fiber $F$, the total monoid space $M$ is the cartesian product
\begin{equation}
M = M_{0} \times \ldots \times M_{i}\times \ldots \times\ldots M_{n}
\end{equation}
of the monoid $M_{i}$ on each $F_{i}\in F$. Because each $F_i, M_i$ can be independent of the other fibers, we can formally express the constraints on heterogenous multivariate datasets in the same manner as univariate datasets.
